An algorithm for the optimization of finite element integration

### 2016-04-07 TOMS_Paper/assembly.tex
Fix last standing issue with plots. Add plotting script
Integrate Paul's suggestions. Still wondering about the title

\begin{abstract}
We present an algorithm for the  optimization of a class of finite element integration loop nests. This algorithm is proven to achieve a locally optimal operation count. In specified circumstances the optimum achieved is global. Extensive numerical experiments demonstrate significant performance improvements over the state of the art in finite element code generation in almost all cases. This validates the algorithm presented here, and illustrate its limitations. 
\end{abstract}

@@

This work was supported by by the
Department of Computing at Imperial College London, the Engineering and
Physical Sciences Research Council [grant number EP/L000407/1], and the Natural
Environment Research Council [grant numbers 
NE/K008951/1 and NE/K006789/1], and by a HiPEAC collaboration
grant. The authors would like to thank Mr. Andrew T.T. McRae,
Dr. Lawrence Mitchell, and Dr. Francis Russell for their invaluable
suggestions and their contribution to the Firedrake project.

@@

As demonstrated by the substantial body of research on the topic, automating the generation of such high performance implementations poses several challenges. This is a result of the complexity inherent in the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals. General-purpose compilers, such as those by \emph{GNU} and \emph{Intel}, fail at exploiting the structure inherent in the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary; we show this is Section~\ref{sec:perf-results}). Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO \cite{PLUTO}, focus on parallelization and optimization for cache locality, treating issues orthogonal to the question of minimising flops. The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation (or synthesizer) systems. \citeN{quadrature1} show how automated code generation can be leveraged to introduce optimizations that a user should not be expected to write ``by hand''. \citeN{FFC-TC} and~\citeN{Francis} employ mathematical reformulations of finite element integration with the aim of minimizing the operation count. In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low level optimizations are analysed. It is also worth mentioning two new new form compilers, UFLACS \cite{Uflacs} and \cite{TSFC}, which particularly target the compile time challenges of the more complex variational forms. The performance evaluation in Section~\ref{sec:perf-results} includes most of these optimization systems.

However, in spite of such a considerable research effort, there is still no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed? In this paper, we formulate an approach that solves this problem for a particular class of forms and provides very good approximations in all other cases (we will define ``local optimality'', which relates operation count with inner loops). In summary, our contributions are as follows:

@@

{\color{red}FIXME: Please replace $nf$ with $\mathrm{nf}$ in all figure captions.}

@@

The sporadic slow-downs (or marginal improvements) exhibited by \texttt{ufls} can be attributed to the presence of sharing in the generated code.

@@

\paragraph{Hyperelasticity}
In the experiments on the hyperelastic model, shown in Figure~\ref{fig:hyperelasticity}, \texttt{cfO2} exhibits the largest gains out of all problem instances considered in this paper. This is a positive result, since it indicates that our transformation algorithm scales well with form complexity. The fact that all code transformation systems (apart from \texttt{tens}) show quite significant speed-ups suggests two points. First, the baseline is highly inefficient. With forms as complex as in the hyperelastic model, a trivial translation of integration routines into code should always be avoided as even the best general-purpose compiler available (the Intel compiler on an Intel platform at maximum optimization level) fails to exploiting the structure inherent in the expressions. Second, the strategy for removing spatial and temporal sharing has a tremendous impact. Sharing elimination as performed by \texttt{cfO2} ensures a critical reduction in operation count, which becomes particularly pronounced for higher values of $q$.

>>>>

\begin{abstract}
We present an algorithm for the optimization of a class of finite element integration loop nests. This algorithm, which exploits fundamental mathematical properties of finite element operators, is proven to achieve a locally optimal operation count. In specified circumstances the optimum achieved is global. Extensive numerical experiments demonstrate significant performance improvements over the state of the art in finite element code generation in almost all cases. This validates the effectiveness of the algorithm presented here, and illustrates its limitations. 
\end{abstract}

@@

This work was supported by by the
Department of Computing at Imperial College London, the Engineering and
Physical Sciences Research Council [grant number EP/L000407/1], and the Natural
Environment Research Council [grant numbers 
NE/K008951/1 and NE/K006789/1], and by a HiPEAC collaboration
grant. The authors would like to thank Dr. Andrew T.T. McRae,
Dr. Lawrence Mitchell, and Dr. Francis Russell for their invaluable
suggestions and their contribution to the Firedrake project.

@@

As demonstrated by the substantial body of research on the topic, automating the generation of such high performance implementations poses several challenges. This is a result of the complexity inherent in the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals. General-purpose compilers, such as those by \emph{GNU} and \emph{Intel}, fail to exploit the structure inherent in the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary; we show this in Section~\ref{sec:perf-results}). Research compilers, for instance those based on polyhedral analysis of loop nests, such as PLUTO \cite{PLUTO}, focus on parallelization and optimization for cache locality, treating issues orthogonal to the question of minimising flops. The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation (or synthesizer) systems. \citeN{quadrature1} show how automated code generation can be leveraged to introduce optimizations that a user should not be expected to write ``by hand''. \citeN{FFC-TC} and~\citeN{Francis} employ mathematical reformulations of finite element integration with the aim of minimizing the operation count. In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low level optimizations are analysed. It is also worth mentioning two new new form compilers, UFLACS \cite{Uflacs} and TSFC \cite{TSFC}, which particularly target the compilation time challenges of the more complex variational forms. The performance evaluation in Section~\ref{sec:perf-results} includes most of these systems.

However, in spite of such a considerable research effort, there is still no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed? In this paper, we formulate an approach that solves this problem for a particular class of forms and provides very good approximations in all other cases. In particular, we will define ``local optimality'', which relates operation count with inner loops. In summary, our contributions are as follows:

@@

@@

The slow-downs (or marginal improvements) seen in a small number of cases exhibited by \texttt{ufls} can be attributed to the presence of sharing in the generated code.

@@

\paragraph{Hyperelasticity}
In the experiments on the hyperelastic model, shown in Figure~\ref{fig:hyperelasticity}, \texttt{cfO2} exhibits the largest gains out of all problem instances considered in this paper. This is a positive result, since it indicates that our transformation algorithm scales well with form complexity. The fact that all code transformation systems (apart from \texttt{tens}) show quite significant speed-ups suggests two points. First, the baseline is highly inefficient. With forms as complex as in the hyperelastic model, a trivial translation of integration routines into code should always be avoided as even the best general-purpose compiler available (the Intel compiler on an Intel platform at maximum optimization level) fails to exploit the structure inherent in the expressions. Second, the strategy for removing spatial and temporal sharing has a tremendous impact. Sharing elimination as performed by \texttt{cfO2} ensures a critical reduction in operation count, which becomes particularly pronounced for higher values of $q$.
